---
title: "01 — Dataset properties (ChEMBL, PubChem, OPERA, RDKit)"
output: html_notebook
---

## Purpose

Builds the main dataset `data/new_drug_dataset.rds` from the drug list.

Run order: **01 -> 02 -> 03**. This notebook is the only place molecular
properties enter the dataset; 02 adds pkasolver pKa and the calculated
LogD/fraction values, 03 builds the separate pKa comparison dataset.

| Block | Source | Columns added |
|---|---|---|
| 1 | `data/Drug_list.xlsx` | `Drug` |
| 2 | ChEMBL molecule API | `ChEMBL_ID, MW, PSA, HBD, HBA, LogP, SMILES` |
| 3 | PubChem PUG REST | fills the above where ChEMBL has no match |
| 4 | manual corrections | fixes for drugs neither API resolves correctly |
| 5 | ChEMBL activity API | `Fu_median, N_fu, Fu_values` |
| 6 | ChEMBL activity API | `PPB_value, Fu_from_PPB, N_ppb, Fu_source, Fu_final` |
| 7 | ChEMBL activity API | `pKa_exp, N_pka, pKa_values` |
| 8 | ChEMBL activity API | `LogD74_exp, N_logd, LogD_values` |
| 9 | OPERA / EPA CompTox | `pKa_acid_pred, pKa_base_pred, LogD74_pred, LogP_opera, Type` |
| 10 | RDKit | `MW_rdkit, PSA_rdkit, HBD_rdkit, HBA_rdkit, LogP_rdkit` + `*_diff_rdkit` |

Block 9 needs a manual step: the `.csv` written in 9a is uploaded to the EPA
CompTox batch-search dashboard, and its result is saved back as
`data/new_opera_output.csv` before 9b is run.

```{r setup, message=FALSE, warning=FALSE}
library(httr2)
library(jsonlite)
library(dplyr)
library(stringr)
library(readr)
library(readxl)
library(rdkitpyr)

MAIN_RDS <- "data/new_drug_dataset.rds"
MAIN_CSV <- "data/new_drug_dataset.csv"

save_main <- function(df) {
  saveRDS(df, MAIN_RDS)
  readr::write_csv(df, MAIN_CSV)
  cat("Saved:", nrow(df), "rows x", ncol(df), "columns\n")
  invisible(df)
}
```

## 1. Drug list

```{r drug-list}
drug_list <- readxl::read_excel("data/Drug_list.xlsx") |>
  dplyr::mutate(
    # Typos in the source spreadsheet
    Drug = dplyr::case_when(
      Drug == "Apixabam" ~ "Apixaban",
      Drug == "Zalepon"  ~ "Zaleplon",
      TRUE ~ Drug
    )
  ) |>
  dplyr::distinct(Drug)

# Names to query the databases under, where the list's label is not what
# ChEMBL/PubChem calls the compound. The Drug column itself never changes, so
# the OPERA join and every downstream file keep working.
lookup_aliases <- c(
  "Ethanol"             = "Alcohol",
  "Dronabinol/THC"      = "Dronabinol",
  "Benzylpenicillin"    = "Penicillin G",
  "Medroxyprogesterone" = "Medroxyprogesterone acetate"
)


leave_unmatched <- c(
  "Thiopentone"     = "duplicate label of Thiopental",
  "Alphamethyldopa" = "duplicate label of Methyldopa"
)

cat("Drugs in list:", nrow(drug_list), "\n")
cat("Lookup aliases:", length(lookup_aliases), "\n")
cat("Left unresolved:", length(leave_unmatched), "\n")

# Every aliased or excluded name must actually be in the list - a typo here
# would silently do nothing.
unknown <- setdiff(c(names(lookup_aliases), names(leave_unmatched)), drug_list$Drug)
if (length(unknown) > 0) {
  warning("Named in lookup_aliases/leave_unmatched but not in the drug list: ",
          paste(unknown, collapse = ", "), call. = FALSE)
}
```

## 2. ChEMBL molecule properties

Exact name match only. The name from the drug list is kept as `Drug` so the
dataset key stays stable regardless of ChEMBL's `pref_name` casing.

```{r chembl-properties, message=TRUE}
fetch_chembl_molecule <- function(drug, query = drug) {
  url <- paste0("https://www.ebi.ac.uk/chembl/api/data/molecule/search.json?q=",
                utils::URLencode(query, reserved = TRUE), "&limit=50")

  parsed <- httr2::request(url) |>
    httr2::req_headers(Accept = "application/json") |>
    httr2::req_perform() |>
    httr2::resp_body_string() |>
    jsonlite::fromJSON(flatten = TRUE)

  molecules <- parsed$molecules
  if (is.null(molecules) || !is.data.frame(molecules) || nrow(molecules) == 0) {
    return(NULL)
  }

  # Exact name match only. Never fall back to the first search result
  match <- molecules |>
    dplyr::filter(tolower(pref_name) == tolower(query)) |>
    dplyr::slice(1)

  if (nrow(match) == 0) return(NULL)

  data.frame(
    Drug      = drug,
    ChEMBL_ID = match$molecule_chembl_id,
    MW        = as.character(match$molecule_properties.full_mwt),
    PSA       = as.character(match$molecule_properties.psa),
    HBD       = as.integer(match$molecule_properties.hbd),
    HBA       = as.integer(match$molecule_properties.hba),
    LogP      = as.character(match$molecule_properties.alogp),
    SMILES    = match$molecule_structures.canonical_smiles,
    Source    = "ChEMBL",
    stringsAsFactors = FALSE
  )
}

empty_row <- function(drug) {
  data.frame(Drug = drug, ChEMBL_ID = NA_character_, MW = NA_character_,
             PSA = NA_character_, HBD = NA_integer_, HBA = NA_integer_,
             LogP = NA_character_, SMILES = NA_character_,
             Source = NA_character_, stringsAsFactors = FALSE)
}

drug_data <- dplyr::bind_rows(lapply(drug_list$Drug, function(drug) {
  if (drug %in% names(leave_unmatched)) {
    message("Left unresolved (", leave_unmatched[[drug]], "): ", drug)
    return(empty_row(drug))
  }

  query <- if (drug %in% names(lookup_aliases)) lookup_aliases[[drug]] else drug

  res <- tryCatch(fetch_chembl_molecule(drug, query),
                  error = function(e) {
                    message("Error for ", drug, ": ", e$message)
                    NULL
                  })
  if (is.null(res)) {
    message("No ChEMBL match: ", drug,
            if (!identical(query, drug)) paste0(" (searched as \"", query, "\")") else "")
    empty_row(drug)
  } else {
    if (!identical(query, drug)) message("Matched ", drug, " as \"", query, "\"")
    res
  }
}))

cat("Matched in ChEMBL:", sum(!is.na(drug_data$ChEMBL_ID)), "/", nrow(drug_data), "\n")
```

## 3. PubChem fallback

Anything still missing MW or SMILES is looked up by name in PubChem.

```{r pubchem-fallback, message=TRUE}
pubchem_by_name <- function(name) {
  url <- paste0("https://pubchem.ncbi.nlm.nih.gov/rest/pug/compound/name/",
                utils::URLencode(name, reserved = TRUE),
                "/property/MolecularWeight,XLogP,TPSA,",
                "HBondDonorCount,HBondAcceptorCount,ConnectivitySMILES/JSON")

  parsed <- httr2::request(url) |>
    httr2::req_perform() |>
    httr2::resp_body_string() |>
    jsonlite::fromJSON(flatten = TRUE)

  props <- parsed$PropertyTable$Properties
  if (is.null(props) || nrow(props) == 0) NULL else props
}

fetch_pubchem <- function(drug) {
  # PubChem's /name/ endpoint rejects a "/" inside the name even encoded, so a
  # combined label like "Dronabinol/THC" is also tried one segment at a time.
  candidates <- unique(trimws(c(drug, strsplit(drug, "/")[[1]])))
  candidates <- candidates[nzchar(candidates)]

  props <- NULL
  matched_name <- NA_character_

  for (nm in candidates) {
    props <- tryCatch(pubchem_by_name(nm), error = function(e) NULL)
    if (!is.null(props)) {
      matched_name <- nm
      break
    }
  }

  if (is.null(props)) return(NULL)
  if (!identical(matched_name, drug)) {
    message("  (matched PubChem under \"", matched_name, "\")")
  }

  data.frame(
    Drug   = drug,
    MW     = as.character(props$MolecularWeight[1]),
    PSA    = as.character(props$TPSA[1]),
    HBD    = as.integer(props$HBondDonorCount[1]),
    HBA    = as.integer(props$HBondAcceptorCount[1]),
    LogP   = as.character(props$XLogP[1]),
    SMILES = as.character(props$ConnectivitySMILES[1]),
    Source = "PubChem",
    stringsAsFactors = FALSE
  )
}

needs_pubchem <- drug_data |>
  dplyr::filter(is.na(MW) | is.na(SMILES)) |>
  dplyr::filter(!Drug %in% names(leave_unmatched)) |>
  dplyr::pull(Drug)

cat("Drugs needing PubChem fallback:", length(needs_pubchem), "\n")
print(needs_pubchem)

for (drug in needs_pubchem) {
  query <- if (drug %in% names(lookup_aliases)) lookup_aliases[[drug]] else drug

  res <- tryCatch(fetch_pubchem(query),
                  error = function(e) {
                    message("PubChem failed for ", drug, ": ", e$message)
                    NULL
                  })
  if (is.null(res)) next

  i <- which(drug_data$Drug == drug)
  for (col in c("MW", "PSA", "HBD", "HBA", "LogP", "SMILES", "Source")) {
    if (is.na(drug_data[[col]][i])) drug_data[[col]][i] <- res[[col]][1]
  }
  message("PubChem filled: ", drug)
}

cat("Still missing MW or SMILES:",
    sum(is.na(drug_data$MW) | is.na(drug_data$SMILES)), "\n")
```

## 4. Manual corrections

Drugs where the API lookups return the wrong record or an incomplete one.
Values are literature/PubChem-confirmed. `NA` means "leave whatever is there".

These are applied **before** the activity queries so the corrected
`ChEMBL_ID` is the one used for Fu, pKa and LogD.

```{r manual-corrections}
manual_fixes <- tibble::tribble(
  ~Drug,              ~ChEMBL_ID,   ~MW,      ~PSA,    ~HBD, ~HBA, ~LogP,  ~SMILES,
  "Amphetamine",      "CHEMBL405",  "135.21", "26.02", 1L,   1L,   "1.58", "CC(N)Cc1ccccc1",
  "Acetaminophen",    "CHEMBL112",  "151.16", "49.33", 2L,   2L,   "1.35", "CC(=O)Nc1ccc(O)cc1",
  # Benzylpenicillin is no longer here: the "Penicillin G" lookup alias now
  # resolves it to CHEMBL29 live, so its values come from the record itself
  # rather than from numbers frozen at the time they were copied in.
  "Ivermectin",       NA,           "875.1",  "170",   3L,   14L,  "4.1",  "CCC(C)C1C(CCC2(O1)CC3CC(O2)CC=C(C(C(C=CC=C4COC5C4(C(C=C(C5O)C)C(=O)O3)O)C)OC6CC(C(C(O6)C)OC7CC(C(C(O7)C)O)OC)OC)C)C",
  "Cyclosporine",     NA,           NA,       "279",   5L,   12L,  "7.5",  NA
)

for (i in seq_len(nrow(manual_fixes))) {
  d <- manual_fixes$Drug[i]
  if (!d %in% drug_data$Drug) {
    message("Manual fix skipped, drug not in list: ", d)
    next
  }
  for (col in setdiff(names(manual_fixes), "Drug")) {
    val <- manual_fixes[[col]][i]
    if (!is.na(val)) drug_data[[col]][drug_data$Drug == d] <- val
  }
  message("Manual fix applied: ", d)
}

drug_data |>
  dplyr::filter(Drug %in% manual_fixes$Drug) |>
  dplyr::select(Drug, ChEMBL_ID, MW, PSA, HBD, HBA, LogP) |>
  print()

# SMILES is required downstream by OPERA, RDKit and pkasolver - a drug without
# one drops out of every prediction column. The deliberately unresolved rows are
# expected to be missing
no_smiles  <- drug_data |> dplyr::filter(is.na(SMILES)) |> dplyr::pull(Drug)
unexpected <- setdiff(no_smiles, names(leave_unmatched))
expected   <- intersect(no_smiles, names(leave_unmatched))

if (length(expected) > 0) {
  cat("No SMILES, as intended:", paste(expected, collapse = ", "), "\n")
}
if (length(unexpected) > 0) {
  warning("No SMILES after all fallbacks - add a lookup alias or a manual fix: ",
          paste(unexpected, collapse = ", "), call. = FALSE)
} else {
  cat("Every other drug has SMILES\n")
}

save_main(drug_data)
```

## 5. Fu — fraction unbound in plasma

Human assays only; animal and tissue assays excluded; values constrained to
`[0, 1]`. Multiple values are summarised as the median and the raw values are
kept in `Fu_values`.

```{r fu-extraction, message=TRUE}
ANIMAL_RX <- "\\brat\\b|\\bmouse\\b|\\bmice\\b|\\banimal\\b|\\brabbit\\b|\\bdog\\b|\\bmonkey\\b"

fetch_activity <- function(chembl_id, standard_type) {
  url <- paste0("https://www.ebi.ac.uk/chembl/api/data/activity.json?molecule_chembl_id=",
                chembl_id, "&standard_type=", utils::URLencode(standard_type), "&limit=100")

  parsed <- httr2::request(url) |>
    httr2::req_headers(Accept = "application/json") |>
    httr2::req_perform() |>
    httr2::resp_body_string() |>
    jsonlite::fromJSON(flatten = TRUE)

  activities <- parsed$activities
  if (is.null(activities) || !is.data.frame(activities) || nrow(activities) == 0) {
    return(NULL)
  }
  activities
}

human_plasma_filter <- function(activities, lo, hi) {
  activities |>
    dplyr::filter(
      (is.na(target_organism) | target_organism == "Homo sapiens") &
        !stringr::str_detect(tolower(assay_description), ANIMAL_RX) &
        stringr::str_detect(tolower(assay_description), "human|plasma") &
        !is.na(standard_value)
    ) |>
    dplyr::mutate(standard_value = as.numeric(standard_value)) |>
    dplyr::filter(standard_value >= lo, standard_value <= hi)
}

fu_results <- dplyr::bind_rows(lapply(seq_len(nrow(drug_data)), function(i) {
  drug      <- drug_data$Drug[i]
  chembl_id <- drug_data$ChEMBL_ID[i]
  blank <- data.frame(Drug = drug, Fu_median = NA_real_, N_fu = 0L,
                      Fu_values = NA_character_, stringsAsFactors = FALSE)

  if (is.na(chembl_id)) return(blank)

  tryCatch({
    activities <- fetch_activity(chembl_id, "Fu")
    if (is.null(activities)) return(blank)

    filtered <- human_plasma_filter(activities, 0, 1)
    if (nrow(filtered) == 0) return(blank)

    data.frame(
      Drug      = drug,
      Fu_median = median(filtered$standard_value, na.rm = TRUE),
      N_fu      = nrow(filtered),
      Fu_values = paste(filtered$standard_value, collapse = ", "),
      stringsAsFactors = FALSE
    )
  }, error = function(e) {
    message("Fu error for ", drug, ": ", e$message)
    blank
  })
}))

drug_data <- drug_data |> dplyr::left_join(fu_results, by = "Drug")

cat("Drugs with Fu:", sum(!is.na(drug_data$Fu_median)), "/", nrow(drug_data), "\n")
```

## 6. PPB fallback → Fu_final

Where no Fu is reported, plasma protein binding (%) is converted with
`Fu = 1 - PPB/100`. `Fu_source` records which route each value came from.

```{r ppb-fallback, message=TRUE}
missing_fu <- drug_data |>
  dplyr::filter(is.na(Fu_median), !is.na(ChEMBL_ID)) |>
  dplyr::select(Drug, ChEMBL_ID)

cat("Drugs missing Fu — checking PPB:", nrow(missing_fu), "\n")

ppb_results <- dplyr::bind_rows(lapply(seq_len(nrow(missing_fu)), function(i) {
  drug      <- missing_fu$Drug[i]
  chembl_id <- missing_fu$ChEMBL_ID[i]
  blank <- data.frame(Drug = drug, PPB_value = NA_real_,
                      Fu_from_PPB = NA_real_, N_ppb = 0L,
                      stringsAsFactors = FALSE)

  tryCatch({
    activities <- fetch_activity(chembl_id, "PPB")
    if (is.null(activities)) return(blank)

    filtered <- human_plasma_filter(activities, 0, 100)
    if (nrow(filtered) == 0) return(blank)

    ppb_median <- median(filtered$standard_value, na.rm = TRUE)
    message(drug, " — PPB ", ppb_median, "% -> Fu ", round(1 - ppb_median / 100, 3))

    data.frame(
      Drug        = drug,
      PPB_value   = ppb_median,
      Fu_from_PPB = 1 - (ppb_median / 100),
      N_ppb       = nrow(filtered),
      stringsAsFactors = FALSE
    )
  }, error = function(e) {
    message("PPB error for ", drug, ": ", e$message)
    blank
  })
}))

drug_data <- drug_data |>
  dplyr::left_join(ppb_results, by = "Drug") |>
  dplyr::mutate(
    Fu_source = dplyr::case_when(
      !is.na(Fu_median)   ~ "ChEMBL_Fu",
      !is.na(Fu_from_PPB) ~ "ChEMBL_PPB",
      TRUE ~ NA_character_
    ),
    Fu_final = dplyr::coalesce(Fu_median, Fu_from_PPB)
  )

cat("Fu from PPB:", sum(drug_data$Fu_source == "ChEMBL_PPB", na.rm = TRUE), "\n")
cat("Fu still missing (literature lookup needed):", sum(is.na(drug_data$Fu_final)), "\n")

drug_data |> dplyr::filter(is.na(Fu_final)) |> dplyr::pull(Drug) |> print()

save_main(drug_data)
```

## 7. Experimental pKa (ChEMBL)


```{r experimental-pka, message=TRUE}
pka_results <- dplyr::bind_rows(lapply(seq_len(nrow(drug_data)), function(i) {
  drug      <- drug_data$Drug[i]
  chembl_id <- drug_data$ChEMBL_ID[i]
  blank <- data.frame(Drug = drug, pKa_exp = NA_real_, N_pka = 0L,
                      pKa_values = NA_character_, stringsAsFactors = FALSE)

  if (is.na(chembl_id)) return(blank)

  tryCatch({
    activities <- fetch_activity(chembl_id, "pKa")
    if (is.null(activities)) return(blank)

    filtered <- activities |>
      dplyr::filter(!is.na(standard_value)) |>
      dplyr::mutate(standard_value = as.numeric(standard_value)) |>
      dplyr::filter(standard_value >= 0, standard_value <= 14)

    if (nrow(filtered) == 0) return(blank)

    data.frame(
      Drug       = drug,
      pKa_exp    = median(filtered$standard_value, na.rm = TRUE),
      N_pka      = nrow(filtered),
      pKa_values = paste(filtered$standard_value, collapse = ", "),
      stringsAsFactors = FALSE
    )
  }, error = function(e) {
    message("pKa error for ", drug, ": ", e$message)
    blank
  })
}))

drug_data <- drug_data |> dplyr::left_join(pka_results, by = "Drug")

cat("Drugs with experimental pKa:", sum(!is.na(drug_data$pKa_exp)), "/", nrow(drug_data), "\n")
```

## 8. Experimental LogD7.4 (ChEMBL)

Restricted to assays whose description mentions pH 7.4 or physiological pH.

```{r experimental-logd, message=TRUE}
logd_results <- dplyr::bind_rows(lapply(seq_len(nrow(drug_data)), function(i) {
  drug      <- drug_data$Drug[i]
  chembl_id <- drug_data$ChEMBL_ID[i]
  blank <- data.frame(Drug = drug, LogD74_exp = NA_real_, N_logd = 0L,
                      LogD_values = NA_character_, stringsAsFactors = FALSE)

  if (is.na(chembl_id)) return(blank)

  tryCatch({
    activities <- fetch_activity(chembl_id, "LogD")
    if (is.null(activities)) return(blank)

    filtered <- activities |>
      dplyr::filter(
        !is.na(standard_value),
        stringr::str_detect(tolower(assay_description), "7\\.4|7\\.40|physiolog")
      ) |>
      dplyr::mutate(standard_value = as.numeric(standard_value))

    if (nrow(filtered) == 0) return(blank)

    data.frame(
      Drug        = drug,
      LogD74_exp  = median(filtered$standard_value, na.rm = TRUE),
      N_logd      = nrow(filtered),
      LogD_values = paste(filtered$standard_value, collapse = ", "),
      stringsAsFactors = FALSE
    )
  }, error = function(e) {
    message("LogD error for ", drug, ": ", e$message)
    blank
  })
}))

drug_data <- drug_data |> dplyr::left_join(logd_results, by = "Drug")

cat("Drugs with experimental LogD7.4:", sum(!is.na(drug_data$LogD74_exp)), "/", nrow(drug_data), "\n")

save_main(drug_data)
```

## 9a. Export OPERA input

Upload the written file to the EPA CompTox batch-search dashboard, request the
OPERA predictions, and save the result as `data/new_opera_output.csv`.

```{r opera-export}
drug_data |>
  dplyr::filter(!is.na(SMILES)) |>
  dplyr::select(Drug, SMILES) |>
  readr::write_csv("data/new_opera_input.csv")

cat("SMILES exported:", sum(!is.na(drug_data$SMILES)), "drugs\n")
```

## 9b. Merge OPERA predictions and assign Type

`Type` classifies each drug from the OPERA predicted pKa values. It is a
descriptive column 

```{r opera-merge}
opera_clean <- readr::read_csv("data/new_opera_output.csv", show_col_types = FALSE) |>
  dplyr::select(
    INPUT,
    pKa_acid_pred = OPERA_PKAA_OPERA_PRED,
    pKa_base_pred = OPERA_PKAB_OPERA_PRED,
    LogD74_pred   = LOGD7.4,
    LogP_opera    = OCTANOL_WATER_PARTITION_LOGP_OPERA_PRED
  ) |>
  dplyr::mutate(dplyr::across(
    c(pKa_acid_pred, pKa_base_pred, LogD74_pred, LogP_opera),
    ~ as.numeric(dplyr::na_if(.x, "N/A"))
  )) |>
  dplyr::distinct(INPUT, .keep_all = TRUE)

drug_data <- drug_data |>
  dplyr::left_join(opera_clean, by = c("Drug" = "INPUT")) |>
  dplyr::mutate(
    Type = dplyr::case_when(
      !is.na(pKa_acid_pred) & !is.na(pKa_base_pred) ~ "AMPHOLYTE",
      !is.na(pKa_acid_pred) & is.na(pKa_base_pred)  ~ "ACID",
      is.na(pKa_acid_pred) & !is.na(pKa_base_pred)  ~ "BASE",
      TRUE ~ "NEUTRAL"
    )
  )

cat("With OPERA pKa acid:", sum(!is.na(drug_data$pKa_acid_pred)), "\n")
cat("With OPERA pKa base:", sum(!is.na(drug_data$pKa_base_pred)), "\n")
cat("With OPERA LogD7.4:", sum(!is.na(drug_data$LogD74_pred)), "\n")
drug_data |> dplyr::count(Type) |> print()

save_main(drug_data)
```

## 10. RDKit descriptors

Two outputs: the curated five descriptors joined into the main dataset with
their ChEMBL differences, and the full descriptor table as its own dataset.

```{r rdkit-descriptors, message=TRUE}
rdkit_curated <- data.frame()
rdkit_full    <- data.frame()

for (i in seq_len(nrow(drug_data))) {
  drug   <- drug_data$Drug[i]
  smiles <- drug_data$SMILES[i]

  blank <- data.frame(Drug = drug, MW_rdkit = NA_real_, PSA_rdkit = NA_real_,
                      HBD_rdkit = NA_real_, HBA_rdkit = NA_real_,
                      LogP_rdkit = NA_real_, stringsAsFactors = FALSE)

  if (is.na(smiles)) {
    rdkit_curated <- dplyr::bind_rows(rdkit_curated, blank)
    rdkit_full    <- dplyr::bind_rows(rdkit_full, data.frame(Drug = drug))
    next
  }

  tryCatch({
    desc <- rdkitpyr::CalculateAllDescriptors(smiles)

    rdkit_curated <- dplyr::bind_rows(rdkit_curated, data.frame(
      Drug       = drug,
      MW_rdkit   = desc$MolWt,
      PSA_rdkit  = desc$TPSA,
      HBD_rdkit  = desc$NumHDonors,
      HBA_rdkit  = desc$NumHAcceptors,
      LogP_rdkit = desc$MolLogP
    ))

    desc$Drug  <- drug
    rdkit_full <- dplyr::bind_rows(rdkit_full, desc)
  }, error = function(e) {
    message("RDKit error for ", drug, ": ", e$message)
    rdkit_curated <<- dplyr::bind_rows(rdkit_curated, blank)
    rdkit_full    <<- dplyr::bind_rows(rdkit_full, data.frame(Drug = drug))
  })

  if (i %% 20 == 0) cat("Processed", i, "/", nrow(drug_data), "\n")
}

drug_data <- drug_data |>
  dplyr::left_join(rdkit_curated, by = "Drug") |>
  dplyr::mutate(
    MW_chembl   = as.numeric(MW),
    PSA_chembl  = as.numeric(PSA),
    LogP_chembl = as.numeric(LogP),
    MW_diff_rdkit   = abs(MW_chembl - MW_rdkit),
    PSA_diff_rdkit  = abs(PSA_chembl - PSA_rdkit),
    HBD_diff_rdkit  = abs(HBD - HBD_rdkit),
    HBA_diff_rdkit  = abs(HBA - HBA_rdkit),
    LogP_diff_rdkit = abs(LogP_chembl - LogP_rdkit)
  )

rdkit_full <- rdkit_full |> dplyr::select(Drug, dplyr::everything())
saveRDS(rdkit_full, "data/rdkit_all_descriptors.rds")
readr::write_csv(rdkit_full, "data/rdkit_all_descriptors.csv")

cat("RDKit descriptors:", sum(!is.na(drug_data$MW_rdkit)), "/", nrow(drug_data), "\n")
cat("Full descriptor table:", ncol(rdkit_full), "columns\n")

save_main(drug_data)
```

## 11. Summary

```{r summary}
n <- nrow(drug_data)

cat("MAIN DATASET :", n, "drugs,", ncol(drug_data), "columns\n\n")
cat("ChEMBL_ID    ", sum(!is.na(drug_data$ChEMBL_ID)),  "/", n, "\n")
cat("SMILES       ", sum(!is.na(drug_data$SMILES)),     "/", n, "\n")
cat("MW           ", sum(!is.na(drug_data$MW)),         "/", n, "\n")
cat("LogP         ", sum(!is.na(drug_data$LogP)),       "/", n, "\n")
cat("Fu_final     ", sum(!is.na(drug_data$Fu_final)),   "/", n, "\n")
cat("pKa_exp      ", sum(!is.na(drug_data$pKa_exp)),    "/", n, "\n")
cat("LogD74_exp   ", sum(!is.na(drug_data$LogD74_exp)), "/", n, "\n")
cat("LogD74_pred  ", sum(!is.na(drug_data$LogD74_pred)),"/", n, "\n")
cat("MW_rdkit     ", sum(!is.na(drug_data$MW_rdkit)),   "/", n, "\n\n")
cat("Next: run 02_pkasolver_logd.Rmd\n")
```