---
title: "02 — pkasolver pKa, LogD7.4 and fraction neutral"
output: html_notebook
---

## Purpose

Adds the pkasolver pKa columns and the values calculated from them to the main
dataset `data/new_drug_dataset.rds`. Run after `01_dataset_properties.Rmd`.

Columns added:

| Column | Meaning |
|---|---|
| `pKa_pkasolver` | every predicted pKa, comma separated, in site order |
| `N_pka_pkasolver` | number of ionisable sites found |
| `Type_pkasolver` | ACID / BASE / OTHER per site, same order |
| `pKa_acid_v1`, `pKa_base_v1` | selection over all sites classified in range 3–11 |
| `pKa_acid_v2`, `pKa_base_v2` | **v2** — same, restricted to SMARTS-identified groups |
| `FG_acid_v2`, `FG_base_v2` | functional group behind each v2 pKa |
| `f_neutral_v2` | fraction neutral at pH 7.4 from the v2 pKa values |
| `fraction_ionized_v2` | `1 - f_neutral_v2` |
| `LogD74_calc_pkasolver_v2` | `LogP_chembl + log10(f_neutral_v2)` |

Acidic sites take the lowest pKa in range, basic sites the highest.
Classification logic lives in `pka_classification.py`.

> **Run this in a fresh R session.** Notebook 01 loads `rdkitpyr`, which
> initialises reticulate against its own Python. Once Python is initialised in
> an R session the environment cannot be switched, so `use_condaenv()` here
> will fail. Restart R after 01 : nothing is lost, since 01 writes everything
> to disk and this notebook reads it back.

```{r setup, message=FALSE, warning=FALSE}
library(reticulate)
library(dplyr)
library(readr)

# Fail with an explanation rather than reticulate's "failed to initialize
# requested version of Python", which does not say what actually went wrong.
if (reticulate::py_available(initialize = FALSE)) {
  stop("Python is already initialised in this R session , most likely by ",
       "rdkitpyr in notebook 01. reticulate cannot switch environments after ",
       "initialisation. Restart R (Session > Restart R) and run this notebook ",
       "first in the fresh session.", call. = FALSE)
}

use_condaenv("pka_prediction", required = TRUE)
source_python("pka_classification.py")

MAIN_RDS <- "data/new_drug_dataset.rds"
MAIN_CSV <- "data/new_drug_dataset.csv"

drug_data <- readRDS(MAIN_RDS)
cat("Loaded:", nrow(drug_data), "rows x", ncol(drug_data), "columns\n")
```

## 1. Run pkasolver

One `calculate_microstate_pka_values` call per molecule. the raw sites and both
selections come back together. Rows that fail return an empty record rather
than dropping out, so `nrow` never changes.

```{r pkasolver-extraction, message=TRUE}
# Python None arrives as NULL; make it an explicit NA of the right type
nz_num <- function(x) if (is.null(x) || length(x) == 0) NA_real_ else as.numeric(x)
nz_chr <- function(x) if (is.null(x) || length(x) == 0) NA_character_ else as.character(x)
nz_int <- function(x) if (is.null(x) || length(x) == 0) 0L else as.integer(x)

PKA_COLS <- c("pKa_pkasolver", "N_pka_pkasolver", "Type_pkasolver",
              "pKa_acid_v1", "pKa_base_v1", "pKa_acid_v2", "pKa_base_v2",
              "FG_acid_v2", "FG_base_v2")

# Drop any previous run so the notebook is safe to re-run
drug_data <- drug_data |> dplyr::select(-dplyr::any_of(PKA_COLS))

pkasolver_results <- dplyr::bind_rows(lapply(seq_len(nrow(drug_data)), function(i) {
  drug   <- drug_data$Drug[i]
  smiles <- drug_data$SMILES[i]

  blank <- data.frame(
    Drug = drug, pKa_pkasolver = NA_character_, N_pka_pkasolver = 0L,
    Type_pkasolver = NA_character_, pKa_acid_v1 = NA_real_, pKa_base_v1 = NA_real_,
    pKa_acid_v2 = NA_real_, pKa_base_v2 = NA_real_,
    FG_acid_v2 = NA_character_, FG_base_v2 = NA_character_,
    stringsAsFactors = FALSE
  )

  if (is.na(smiles)) return(blank)

  res <- tryCatch(process_drug_full(smiles),
                  error = function(e) {
                    message("pkasolver error for ", drug, ": ", e$message)
                    NULL
                  })
  if (is.null(res)) return(blank)

  if (i %% 20 == 0) cat("Processed", i, "/", nrow(drug_data), "\n")

  data.frame(
    Drug            = drug,
    pKa_pkasolver   = nz_chr(res$pKa_pkasolver),
    N_pka_pkasolver = nz_int(res$N_pka_pkasolver),
    Type_pkasolver  = nz_chr(res$Type_pkasolver),
    pKa_acid_v1     = nz_num(res$pKa_acid_v1),
    pKa_base_v1     = nz_num(res$pKa_base_v1),
    pKa_acid_v2     = nz_num(res$pKa_acid_v2),
    pKa_base_v2     = nz_num(res$pKa_base_v2),
    FG_acid_v2      = nz_chr(res$FG_acid_v2),
    FG_base_v2      = nz_chr(res$FG_base_v2),
    stringsAsFactors = FALSE
  )
}))

drug_data <- drug_data |> dplyr::left_join(pkasolver_results, by = "Drug")

cat("\nPKASOLVER SUMMARY\n")
cat("Sites found:     ", sum(drug_data$N_pka_pkasolver > 0), "/", nrow(drug_data), "\n")
cat("v1 acid / base:  ", sum(!is.na(drug_data$pKa_acid_v1)), "/",
                         sum(!is.na(drug_data$pKa_base_v1)), "\n")
cat("v2 acid / base:  ", sum(!is.na(drug_data$pKa_acid_v2)), "/",
                         sum(!is.na(drug_data$pKa_base_v2)), "\n")
```

## 2. Silent-failure check

Sites found but neither v1 value set means the classification returned nothing
usable — the pattern that hid the earlier `process_drug_batch` bug. This should
print zero rows.

```{r silent-failure-check}
suspicious <- drug_data |>
  dplyr::filter(N_pka_pkasolver > 0, is.na(pKa_acid_v1), is.na(pKa_base_v1)) |>
  dplyr::select(Drug, SMILES, N_pka_pkasolver, Type_pkasolver)

cat("Drugs with sites but no v1 classification:", nrow(suspicious), "\n")
if (nrow(suspicious) > 0) print(as.data.frame(suspicious))
```

## 3. LogD7.4 and fraction neutral

Single combined formula across both ionisable sites:

```
f_neutral = 1 / (1 + 10^(pH - pKa_acid) + 10^(pKa_base - pH))
```

A missing acid or base pKa contributes 0 to the denominator. A drug with
neither is neutral, so `f_neutral = 1` and LogD equals LogP.

```{r logd-calculation}
pH <- 7.4

drug_data <- drug_data |>
  dplyr::mutate(
    # A drug with no SMILES was never assessed at all, which is not the same as
    # a drug pkasolver assessed and found no ionisable site in. Without this
    # guard both pKa terms are absent, f_neutral comes out as exactly 1 and
    # fraction_ionized as 0 - reading as "confirmed neutral" when the truth is
    # "unknown". Rows that were assessed and genuinely have no site keep f = 1,
    # which is correct for them.
    .assessed = !is.na(SMILES),
    f_neutral_v2 = ifelse(
      .assessed,
      1 / (1 +
        ifelse(!is.na(pKa_acid_v2), 10^(pH - pKa_acid_v2), 0) +
        ifelse(!is.na(pKa_base_v2), 10^(pKa_base_v2 - pH), 0)),
      NA_real_
    ),
    fraction_ionized_v2      = 1 - f_neutral_v2,
    LogD74_calc_pkasolver_v2 = LogP_chembl + log10(f_neutral_v2)
  ) |>
  dplyr::select(-.assessed)

not_assessed <- drug_data |> dplyr::filter(is.na(SMILES)) |> dplyr::pull(Drug)
if (length(not_assessed) > 0) {
  cat("Not assessed (no SMILES), left as NA:",
      paste(not_assessed, collapse = ", "), "\n\n")
}

cat("f_neutral_v2:             ", sum(!is.na(drug_data$f_neutral_v2)), "/", nrow(drug_data), "\n")
cat("fraction_ionized_v2:      ", sum(!is.na(drug_data$fraction_ionized_v2)), "/", nrow(drug_data), "\n")
cat("LogD74_calc_pkasolver_v2: ", sum(!is.na(drug_data$LogD74_calc_pkasolver_v2)), "/", nrow(drug_data), "\n\n")

cat("fraction_ionized_v2 distribution:\n")
print(summary(drug_data$fraction_ionized_v2))
```

## 4. Save

```{r save}
saveRDS(drug_data, MAIN_RDS)
readr::write_csv(drug_data, MAIN_CSV)

cat("MAIN DATASET COMPLETE :", nrow(drug_data), "rows x", ncol(drug_data), "columns\n")
cat("Next: run 03_pka_comparison.Rmd for the pKa method comparison dataset\n")
```