---
title: "03 — pKa method comparison dataset"
output: html_notebook
---

## Purpose

Builds the smaller side dataset `data/pka_comparison.rds`, holding the acid and
base pKa for each drug from four sources:

| Source | Where it comes from |
|---|---|
| pkasolver v2 | already in the main dataset (notebook 02) |
| OPERA | already in the main dataset (notebook 01) |
| MolGpKa | fetched here from the xundrug API |
| Uni-pKa | `unipka_predict.py`, run separately |

Nothing here is written back into the main dataset . it is a comparison table
only. Run after `01` and `02`.

All four sources are filtered to pKa 3–11; acid takes the lowest value in
range, base the highest.

```{r setup, message=FALSE, warning=FALSE}
library(httr2)
library(jsonlite)
library(dplyr)
library(readr)

MAIN_RDS  <- "data/new_drug_dataset.rds"
UNIPKA_CSV <- "data/unipka_macro_pka.csv"
OUT_RDS   <- "data/pka_comparison.rds"
OUT_CSV   <- "data/pka_comparison.csv"

PKA_MIN <- 3
PKA_MAX <- 11

drug_data <- readRDS(MAIN_RDS)
cat("Main dataset:", nrow(drug_data), "drugs\n")
```

## 1. MolGpKa

Undocumented batch endpoint; one molecule per request, so the loop is paced to
avoid rate limiting and failures are retried once with a longer pause.

```{r molgpka-function}
MOLGPKA_URL   <- "http://xundrug.cn:5001/modules/upload0/"
MOLGPKA_TOKEN <- "O05DriqqQLlry9kmpCwms2IJLC0MuLQ7"

get_molgpka <- function(smiles) {
  tryCatch({
    temp_file <- tempfile(fileext = ".txt")
    on.exit(unlink(temp_file), add = TRUE)
    writeLines(smiles, temp_file, sep = "")

    result <- httr2::request(MOLGPKA_URL) |>
      httr2::req_headers(token = MOLGPKA_TOKEN) |>
      httr2::req_body_multipart(Smiles = curl::form_file(temp_file, name = "tmg")) |>
      httr2::req_perform() |>
      httr2::resp_body_json()

    list(acid = result$gen_datas$Acid, base = result$gen_datas$Base)
  }, error = function(e) {
    message("MolGpKa error: ", e$message)
    list(acid = NULL, base = NULL)
  })
}

# Turn one API response into a single row
molgpka_row <- function(drug, result) {
  acid_nums <- if (length(result$acid) > 0) as.numeric(unlist(result$acid)) else numeric(0)
  base_nums <- if (length(result$base) > 0) as.numeric(unlist(result$base)) else numeric(0)

  acid_in_range <- acid_nums[acid_nums >= PKA_MIN & acid_nums <= PKA_MAX]
  base_in_range <- base_nums[base_nums >= PKA_MIN & base_nums <= PKA_MAX]

  data.frame(
    Drug                 = drug,
    molgpka_acid_raw     = if (length(acid_nums) > 0) paste(acid_nums, collapse = ", ") else NA_character_,
    molgpka_base_raw     = if (length(base_nums) > 0) paste(base_nums, collapse = ", ") else NA_character_,
    molgpka_acid_N       = length(acid_nums),
    molgpka_base_N       = length(base_nums),
    molgpka_pKa_acid     = if (length(acid_in_range) > 0) min(acid_in_range) else NA_real_,
    molgpka_pKa_base     = if (length(base_in_range) > 0) max(base_in_range) else NA_real_,
    stringsAsFactors = FALSE
  )
}
```

```{r molgpka-extraction, message=TRUE}
molgpka_results <- dplyr::bind_rows(lapply(seq_len(nrow(drug_data)), function(i) {
  drug   <- drug_data$Drug[i]
  smiles <- drug_data$SMILES[i]

  if (is.na(smiles)) return(molgpka_row(drug, list(acid = NULL, base = NULL)))

  row <- molgpka_row(drug, get_molgpka(smiles))
  message(i, "/", nrow(drug_data), " - ", drug)
  Sys.sleep(1)
  row
}))

cat("MolGpKa acid (raw):", sum(!is.na(molgpka_results$molgpka_acid_raw)), "\n")
cat("MolGpKa base (raw):", sum(!is.na(molgpka_results$molgpka_base_raw)), "\n")
```

```{r molgpka-retry, message=TRUE}
# Drugs that returned nothing at all are usually rate limiting, not real
# absences - retry once at a slower pace.
failed <- molgpka_results |>
  dplyr::filter(is.na(molgpka_acid_raw), is.na(molgpka_base_raw)) |>
  dplyr::pull(Drug)

cat("Retrying:", length(failed), "drugs\n")

for (drug in failed) {
  smiles <- drug_data$SMILES[drug_data$Drug == drug]
  if (length(smiles) == 0 || is.na(smiles)) next

  row <- molgpka_row(drug, get_molgpka(smiles))
  if (!is.na(row$molgpka_acid_raw) || !is.na(row$molgpka_base_raw)) {
    molgpka_results[molgpka_results$Drug == drug, names(row)] <- row
    message(drug, " - recovered")
  }
  Sys.sleep(5)
}

cat("Still empty after retry:",
    sum(is.na(molgpka_results$molgpka_acid_raw) & is.na(molgpka_results$molgpka_base_raw)), "\n")
```

## 2. Uni-pKa

Produced by `unipka_predict.py`, which runs in the `unipka` environment (not
the `pka_prediction` one)  If the file is not there yet, the Uni-pKa columns come through as NA.

```{r unipka-load}
if (file.exists(UNIPKA_CSV)) {
  unipka_results <- readr::read_csv(UNIPKA_CSV, show_col_types = FALSE) |>
    dplyr::select(
      Drug,
      unipka_pKa_acid = unipka_acidic_macro_pKa,
      unipka_pKa_base = unipka_basic_macro_pKa
    )
  cat("Uni-pKa acid:", sum(!is.na(unipka_results$unipka_pKa_acid)), "\n")
  cat("Uni-pKa base:", sum(!is.na(unipka_results$unipka_pKa_base)), "\n")
} else {
  warning("Uni-pKa file not found - run unipka_predict.py. Columns will be NA.")
  unipka_results <- data.frame(
    Drug = drug_data$Drug,
    unipka_pKa_acid = NA_real_,
    unipka_pKa_base = NA_real_,
    stringsAsFactors = FALSE
  )
}
```

## 3. Assemble the comparison dataset

```{r build-comparison}
pka_comparison <- drug_data |>
  dplyr::select(
    Drug, SMILES, Type,
    pkasolver_pKa_acid = pKa_acid_v2,
    pkasolver_pKa_base = pKa_base_v2,
    pkasolver_FG_acid  = FG_acid_v2,
    pkasolver_FG_base  = FG_base_v2,
    opera_pKa_acid     = pKa_acid_pred,
    opera_pKa_base     = pKa_base_pred,
    pKa_exp
  ) |>
  dplyr::left_join(molgpka_results, by = "Drug") |>
  dplyr::left_join(unipka_results, by = "Drug") |>
  dplyr::select(
    Drug, SMILES, Type, pKa_exp,
    pkasolver_pKa_acid, opera_pKa_acid, molgpka_pKa_acid, unipka_pKa_acid,
    pkasolver_pKa_base, opera_pKa_base, molgpka_pKa_base, unipka_pKa_base,
    pkasolver_FG_acid, pkasolver_FG_base,
    molgpka_acid_raw, molgpka_base_raw, molgpka_acid_N, molgpka_base_N
  )

saveRDS(pka_comparison, OUT_RDS)
readr::write_csv(pka_comparison, OUT_CSV)

cat("Saved:", nrow(pka_comparison), "rows x", ncol(pka_comparison), "columns\n")
```

## 4. Coverage

How many drugs each method returns an in-range pKa for.

```{r coverage}
coverage <- data.frame(
  Method = c("pkasolver v2", "OPERA", "MolGpKa", "Uni-pKa"),
  N_acid = c(
    sum(!is.na(pka_comparison$pkasolver_pKa_acid)),
    sum(!is.na(pka_comparison$opera_pKa_acid)),
    sum(!is.na(pka_comparison$molgpka_pKa_acid)),
    sum(!is.na(pka_comparison$unipka_pKa_acid))
  ),
  N_base = c(
    sum(!is.na(pka_comparison$pkasolver_pKa_base)),
    sum(!is.na(pka_comparison$opera_pKa_base)),
    sum(!is.na(pka_comparison$molgpka_pKa_base)),
    sum(!is.na(pka_comparison$unipka_pKa_base))
  )
)
coverage$N_drugs <- nrow(pka_comparison)

print(coverage)
```