---
title: "Dataset creation"
output: html_notebook
---



## Overview

This notebook curates a dataset of drugs for use in milk to plasma (MP) ratio model training. For each drug, the following variables are extracted or calculated:

- **MW, PSA, HBD, HBA, SMILES** : extracted from ChEMBL molecule API
- **Fu** : extracted from ChEMBL activity API (human plasma only)
- **LogP** : extracted from ChEMBL molecule API (alogp field)
- **pKa** : extracted from ChEMBL activity API (experimental values)
- **LogD7.4** : three columns: experimental (ChEMBL), calculated (from LogP + pKa), predicted (SMILES-based model)

PubChem is used as a fallback if a drug is not found in ChEMBL.



## Setup
```{r}
library(httr2)
library(jsonlite)
library(dplyr)
library(readxl)
```



```{r drug-list}
library(readxl)

drug_names <- readxl::read_excel("data/drug.xlsx") |>
  dplyr::pull(Drug)

cat("Total drugs:", length(drug_names))
```


## Step 1: Extract Molecule Properties from ChEMBL
Extracts MW, PSA, HBD, HBA, LogP, and SMILES from the ChEMBL molecule API.
Filters for exact name matches only. Duplicates are removed after joining.

```{r molecule-extraction, message=TRUE}
molecule_results <- data.frame()

for (drug in drug_names) {
  message(paste("Fetching molecule data for:", drug))
  
  tryCatch({
    url <- paste0("https://www.ebi.ac.uk/chembl/api/data/molecule/search.json?q=",
                  utils::URLencode(drug), "&limit=50")
    
    req <- httr2::request(url) |> httr2::req_headers(Accept = "application/json")
    response <- httr2::req_perform(req)
    parsed <- jsonlite::fromJSON(httr2::resp_body_string(response), flatten = TRUE)
    
    molecules <- parsed$molecules
    
    if (is.null(molecules) || !is.data.frame(molecules) || nrow(molecules) == 0) {
      message(paste("No results found for:", drug))
    } else {
      match <- molecules |>
        dplyr::filter(tolower(pref_name) == tolower(drug)) |>
        dplyr::select(
          pref_name,
          molecule_chembl_id,
          molecule_properties.full_mwt,
          molecule_properties.psa,
          molecule_properties.hbd,
          molecule_properties.hba,
          molecule_properties.alogp,
          molecule_structures.canonical_smiles
        ) |>
        dplyr::slice(1)
      
      if (nrow(match) == 0) {
        message(paste("No exact match for:", drug))
      } else {
        match$Drug_Input <- drug
        molecule_results <<- dplyr::bind_rows(molecule_results, match)
      }
    }
    
  }, error = function(e) {
    message(paste("Error processing", drug, ":", e$message))
  })
}

molecule_results <- molecule_results |>
  dplyr::distinct(pref_name, molecule_chembl_id, .keep_all = TRUE)

cat("Drugs successfully matched:", nrow(molecule_results), "of", length(drug_names))
```

## Step 2: Extract Fu from ChEMBL Activity API
Extracts unbound fraction in plasma (Fu) from ChEMBL activity data.

Filtering logic:

- Human assays only: target_organism is NA or Homo sapiens
- Plasma measurements only: assay_description contains "plasma" or "human"
- Tissue measurements excluded
- Animal keywords excluded (using word boundaries to avoid substring matches)
- Fu values of exactly 0 or 1 are flagged for manual literature review



```{r fu-extraction, message=TRUE}
fu_results <- data.frame()

for (chembl_id in molecule_results$molecule_chembl_id) {
  message(paste("Fetching Fu data for:", chembl_id))
  
  tryCatch({
    url <- paste0("https://www.ebi.ac.uk/chembl/api/data/activity.json?molecule_chembl_id=",
                  chembl_id, "&standard_type=Fu&limit=100")
    
    req <- httr2::request(url) |> httr2::req_headers(Accept = "application/json")
    response <- httr2::req_perform(req)
    parsed <- jsonlite::fromJSON(httr2::resp_body_string(response), flatten = TRUE)
    
    activities <- parsed$activities
    
    if (is.null(activities) || !is.data.frame(activities) || nrow(activities) == 0) {
      fu_results <<- dplyr::bind_rows(fu_results,
        data.frame(molecule_chembl_id = chembl_id,
                   Fu_median = NA, N_fu = 0,
                   Fu_values = NA, Fu_flag = NA))
    } else {
      filtered <- activities |>
        dplyr::filter(
          (is.na(target_organism) | target_organism == "Homo sapiens") &
          !stringr::str_detect(tolower(assay_description),
                               "\\brat\\b|\\bmouse\\b|\\bmice\\b|\\banimal\\b|\\brabbit\\b|\\bdog\\b|\\bmonkey\\b") &
          !stringr::str_detect(tolower(assay_description), "tissue") &
          stringr::str_detect(tolower(assay_description), "plasma|human") &
          !is.na(standard_value)
        ) |>
        dplyr::mutate(standard_value = as.numeric(standard_value)) |>
        dplyr::filter(standard_value >= 0, standard_value <= 1)
      
      if (nrow(filtered) == 0) {
        fu_results <<- dplyr::bind_rows(fu_results,
          data.frame(molecule_chembl_id = chembl_id,
                     Fu_median = NA, N_fu = 0,
                     Fu_values = NA, Fu_flag = NA))
      } else {
        fu_summary <- filtered |>
          dplyr::summarise(
            Fu_median = median(standard_value, na.rm = TRUE),
            N_fu = dplyr::n(),
            Fu_values = paste(standard_value, collapse = ", "),
            Fu_flag = ifelse(any(standard_value == 0 | standard_value == 1),
                             "FLAG: contains 0 or 1", NA)
          ) |>
          dplyr::mutate(molecule_chembl_id = chembl_id)
        
        fu_results <<- dplyr::bind_rows(fu_results, fu_summary)
      }
    }
    
  }, error = function(e) {
    message(paste("Error processing", chembl_id, ":", e$message))
    fu_results <<- dplyr::bind_rows(fu_results,
      data.frame(molecule_chembl_id = chembl_id,
                 Fu_median = NA, N_fu = 0,
                 Fu_values = NA, Fu_flag = NA))
  })
}
```


## Step 3: Combine and Review
```{r combine}
final_dataset <- molecule_results |>
  dplyr::left_join(fu_results, by = "molecule_chembl_id") |>
  dplyr::rename(
    Drug      = pref_name,
    ChEMBL_ID = molecule_chembl_id,
    MW        = molecule_properties.full_mwt,
    PSA       = molecule_properties.psa,
    HBD       = molecule_properties.hbd,
    HBA       = molecule_properties.hba,
    LogP      = molecule_properties.alogp,
    SMILES    = molecule_structures.canonical_smiles
  ) |>
  dplyr::select(Drug, ChEMBL_ID, MW, PSA, HBD, HBA, LogP,
                Fu_median, N_fu, Fu_values, Fu_flag, SMILES)

final_dataset
```

## Step 4 : Quality Check

```{r quality-check}
cat("DATASET SUMMARY\n")
cat("Total drugs:", nrow(final_dataset), "\n")
cat("Drugs with Fu:", sum(!is.na(final_dataset$Fu_median)), "\n")
cat("Drugs missing Fu (need literature lookup):", sum(is.na(final_dataset$Fu_median)), "\n")
cat("Drugs with flagged Fu (need literature review):", sum(!is.na(final_dataset$Fu_flag)), "\n")
cat("Drugs missing MW (need investigation):", sum(is.na(final_dataset$MW)), "\n")

cat("\nDRUGS MISSING Fu\n")
final_dataset |>
  dplyr::filter(is.na(Fu_median)) |>
  dplyr::select(Drug, ChEMBL_ID) |>
  print()

cat("\nFLAGGED Fu VALUES\n")
final_dataset |>
  dplyr::filter(!is.na(Fu_flag)) |>
  dplyr::select(Drug, Fu_median, N_fu, Fu_values, Fu_flag) |>
  print()

cat("\nDRUGS MISSING PROPERTIES\n")
final_dataset |>
  dplyr::filter(is.na(MW)) |>
  dplyr::select(Drug, ChEMBL_ID) |>
  print()
```

```{r unmatched-drugs}
unmatched <- drug_names[!tolower(drug_names) %in% tolower(final_dataset$Drug)]
cat("Unmatched drugs:", length(unmatched), "\n\n")
print(unmatched)
```

```{r check-names}
uncertain <- c("Estradiol", "Ethanol", "Benzylpenicillin", 
               "Methotrexate", "Azapropazone", "Medroxyprogesterone")

for (drug in uncertain) {
  url <- paste0("https://www.ebi.ac.uk/chembl/api/data/molecule/search.json?q=",
                utils::URLencode(drug), "&limit=5")
  
  req <- httr2::request(url) |> httr2::req_headers(Accept = "application/json")
  response <- httr2::req_perform(req)
  parsed <- jsonlite::fromJSON(httr2::resp_body_string(response), flatten = TRUE)
  
  if (!is.null(parsed$molecules) && nrow(parsed$molecules) > 0) {
    cat("\n", drug, "— top matches in ChEMBL:\n")
    print(parsed$molecules$pref_name[1:min(3, nrow(parsed$molecules))])
  } else {
    cat("\n", drug, "— no results found\n")
  }
}
```

```{r pubchem-fallback}
missing_props <- final_dataset |>
  dplyr::filter(is.na(MW)) |>
  dplyr::pull(Drug)

cat("Drugs needing PubChem fallback:", length(missing_props), "\n")
print(missing_props)

manual_cids <- list(
  "IVERMECTIN" = 6321424
)

pubchem_results <- data.frame()

for (drug in missing_props) {
  message(paste("Fetching PubChem data for:", drug))
  
  if (toupper(drug) %in% names(manual_cids)) {
    cid <- manual_cids[[toupper(drug)]]
    url <- paste0("https://pubchem.ncbi.nlm.nih.gov/rest/pug/compound/cid/",
                  cid,
                  "/property/MolecularWeight,XLogP,TPSA,HBondDonorCount,HBondAcceptorCount,ConnectivitySMILES/JSON")
  } else {
    url <- paste0("https://pubchem.ncbi.nlm.nih.gov/rest/pug/compound/name/",
                  utils::URLencode(drug),
                  "/property/MolecularWeight,XLogP,TPSA,HBondDonorCount,HBondAcceptorCount,ConnectivitySMILES/JSON")
  }
  
  req <- httr2::request(url)
  response <- httr2::req_perform(req)
  parsed <- jsonlite::fromJSON(httr2::resp_body_string(response), flatten = TRUE)
  props <- parsed$PropertyTable$Properties
  
  new_row <- data.frame(
    Drug   = toupper(drug),
    MW     = as.character(props$MolecularWeight),
    PSA    = as.character(props$TPSA),
    HBD    = as.integer(props$HBondDonorCount),
    HBA    = as.integer(props$HBondAcceptorCount),
    LogP   = as.character(props$XLogP),
    SMILES = as.character(props$ConnectivitySMILES),
    Source = "PubChem"
  )
  
  pubchem_results <- dplyr::bind_rows(pubchem_results, new_row)
  
  final_dataset <<- final_dataset |>
    dplyr::mutate(
      MW     = ifelse(Drug == toupper(drug), new_row$MW, MW),
      PSA    = ifelse(Drug == toupper(drug), new_row$PSA, PSA),
      HBD    = ifelse(Drug == toupper(drug), new_row$HBD, HBD),
      HBA    = ifelse(Drug == toupper(drug), new_row$HBA, HBA),
      LogP   = ifelse(Drug == toupper(drug), new_row$LogP, LogP),
      SMILES = ifelse(Drug == toupper(drug), new_row$SMILES, SMILES)
    )
  
  message(paste("Updated", drug, "with PubChem data"))
}

cat("\nPubChem fallback complete\n")
print(pubchem_results)
```
```{r check-ivermectin}
final_dataset |> dplyr::filter(Drug == "IVERMECTIN")
```

```{r check-fu-ppb}
chembl_id <- "CHEMBL578"

# Try alternate standard_types used for plasma protein binding / Fu
types_to_try <- c("Fu", "fu", "PPB", "Plasma protein binding", 
                  "Fraction unbound", "Unbound fraction", "fup")

results_list <- lapply(types_to_try, function(stype) {
  url <- paste0("https://www.ebi.ac.uk/chembl/api/data/activity.json?molecule_chembl_id=",
                chembl_id, "&standard_type=", utils::URLencode(stype), "&limit=100")
  
  tryCatch({
    req <- httr2::request(url) |> httr2::req_headers(Accept = "application/json")
    response <- httr2::req_perform(req)
    parsed <- jsonlite::fromJSON(httr2::resp_body_string(response), flatten = TRUE)
    
    acts <- parsed$activities
    if (is.data.frame(acts) && nrow(acts) > 0) {
      message(stype, ": ", nrow(acts), " result(s) found")
      acts$queried_type <- stype
      return(acts)
    } else {
      message(stype, ": no results")
      return(NULL)
    }
  }, error = function(e) {
    message(stype, ": error - ", e$message)
    return(NULL)
  })
})

# Combine whatever came back
hits <- dplyr::bind_rows(results_list)

if (nrow(hits) > 0) {
  hits |>
  dplyr::mutate(dplyr::across(where(is.list),
    ~ sapply(.x, function(v) paste(unlist(v), collapse = "; ")))) |>
  dplyr::select(queried_type, standard_type, standard_value,
                standard_units, assay_description) |>
  as.data.frame() |>   # force out of tibble to avoid na.print issue
  print()
} else {
  message("No PPB/Fu data found under any standard_type for ", chembl_id)
}
```

```{r}
final_dataset
```

```{r check-fu-duloxetine}
url <- "https://www.ebi.ac.uk/chembl/api/data/activity.json?molecule_chembl_id=CHEMBL1175&standard_type=Fu&limit=100"

req <- httr2::request(url) |> httr2::req_headers(Accept = "application/json")
response <- httr2::req_perform(req)
parsed <- jsonlite::fromJSON(httr2::resp_body_string(response), flatten = TRUE)

activities <- parsed$activities

if (is.null(activities) || !is.data.frame(activities) || nrow(activities) == 0) {
  cat("No Fu data found for Duloxetine\n")
} else {
  activities |>
    dplyr::select(standard_value, target_organism, assay_description) |>
    print()
}
```

```{r check-fu-missing}
# Check Olanzapine and Escitalopram
test_ids <- list(
  "Olanzapine" = "CHEMBL715",
  "Escitalopram" = "CHEMBL1508"
)

for (name in names(test_ids)) {
  url <- paste0("https://www.ebi.ac.uk/chembl/api/data/activity.json?molecule_chembl_id=",
                test_ids[[name]], "&standard_type=Fu&limit=100")
  
  req <- httr2::request(url) |> httr2::req_headers(Accept = "application/json")
  response <- httr2::req_perform(req)
  parsed <- jsonlite::fromJSON(httr2::resp_body_string(response), flatten = TRUE)
  activities <- parsed$activities
  
  if (is.null(activities) || !is.data.frame(activities) || nrow(activities) == 0) {
    cat(name, "— No Fu data in ChEMBL at all\n")
  } else {
    cat(name, "— Fu data found:\n")
    activities |>
      dplyr::select(standard_value, target_organism, assay_description) |>
      print()
  }
}
```

```{r fu-ppb-fallback, message=TRUE}

final_dataset <- final_dataset |>
  dplyr::select(-dplyr::any_of(c("PPB_value", "Fu_from_PPB", "N_ppb", "Fu_source", "Fu_final")))


# Getting drugs still missing Fu
missing_fu <- final_dataset |>
  dplyr::filter(is.na(Fu_median)) |>
  dplyr::select(Drug, ChEMBL_ID)

cat("Drugs missing Fu — checking for PPB data:", nrow(missing_fu), "\n\n")

ppb_results <- data.frame()

for (i in 1:nrow(missing_fu)) {
  drug <- missing_fu$Drug[i]
  chembl_id <- missing_fu$ChEMBL_ID[i]
  
  tryCatch({
    url <- paste0("https://www.ebi.ac.uk/chembl/api/data/activity.json?molecule_chembl_id=",
                  chembl_id, "&standard_type=PPB&limit=100")
    
    req <- httr2::request(url) |> httr2::req_headers(Accept = "application/json")
    response <- httr2::req_perform(req)
    parsed <- jsonlite::fromJSON(httr2::resp_body_string(response), flatten = TRUE)
    activities <- parsed$activities
    
    if (is.null(activities) || !is.data.frame(activities) || nrow(activities) == 0) {
      ppb_results <- dplyr::bind_rows(ppb_results,
        data.frame(Drug = drug, ChEMBL_ID = chembl_id,
                   PPB_value = NA, Fu_from_PPB = NA, N_ppb = 0))
    } else {
      # Applying same human filter as Fu
      filtered <- activities |>
        dplyr::filter(
          (is.na(target_organism) | target_organism == "Homo sapiens") &
          !stringr::str_detect(tolower(assay_description),
                               "\\brat\\b|\\bmouse\\b|\\bmice\\b|\\banimal\\b|\\brabbit\\b|\\bdog\\b|\\bmonkey\\b") &
          stringr::str_detect(tolower(assay_description), "human|plasma") &
          !is.na(standard_value)
        ) |>
        dplyr::mutate(standard_value = as.numeric(standard_value)) |>
        dplyr::filter(standard_value >= 0, standard_value <= 100)
      
      if (nrow(filtered) == 0) {
        ppb_results <- dplyr::bind_rows(ppb_results,
          data.frame(Drug = drug, ChEMBL_ID = chembl_id,
                     PPB_value = NA, Fu_from_PPB = NA, N_ppb = 0))
      } else {
        ppb_median <- median(filtered$standard_value, na.rm = TRUE)
        fu_calculated <- 1 - (ppb_median / 100)
        
        ppb_results <- dplyr::bind_rows(ppb_results,
          data.frame(Drug = drug, ChEMBL_ID = chembl_id,
                     PPB_value = ppb_median,
                     Fu_from_PPB = fu_calculated,
                     N_ppb = nrow(filtered)))
        
        message(paste(drug, "— PPB found:", ppb_median, "% → Fu =", round(fu_calculated, 3)))
      }
    }
    
  }, error = function(e) {
    message(paste("Error for", drug, ":", e$message))
    ppb_results <<- dplyr::bind_rows(ppb_results,
      data.frame(Drug = drug, ChEMBL_ID = chembl_id,
                 PPB_value = NA, Fu_from_PPB = NA, N_ppb = 0))
  })
}

# Update final_dataset with PPB derived Fu
final_dataset <- final_dataset |>
  dplyr::left_join(ppb_results |> dplyr::select(ChEMBL_ID, PPB_value, Fu_from_PPB, N_ppb),
                   by = "ChEMBL_ID") |>
  dplyr::mutate(
    Fu_source = dplyr::case_when(
      !is.na(Fu_median) ~ "ChEMBL_Fu",
      !is.na(Fu_from_PPB) ~ "ChEMBL_PPB",
      TRUE ~ NA_character_
    ),
    Fu_final = dplyr::coalesce(Fu_median, Fu_from_PPB)
  )

cat("\nPPB FALLBACK SUMMARY\n")
cat("Drugs with Fu from PPB:", sum(!is.na(final_dataset$Fu_from_PPB) & is.na(final_dataset$Fu_median)), "\n")
cat("Drugs still missing Fu:", sum(is.na(final_dataset$Fu_final)), "\n")

cat("\nSTILL MISSING Fu (need literature review)\n")
final_dataset |>
  dplyr::filter(is.na(Fu_final)) |>
  dplyr::select(Drug, ChEMBL_ID) |>
  print()
```
```{r save-dataset}
# Save final dataset
saveRDS(final_dataset, "data/final_dataset.rds")
readr::write_csv(final_dataset, "data/final_dataset.csv")
cat("Dataset saved successfully\n")
cat("Rows:", nrow(final_dataset), "\n")
cat("Columns:", ncol(final_dataset), "\n")
```



