---
title: "Dataset creation"
output: html_notebook
---



## Overview

This notebook curates a dataset of drugs for use in milk to plasma (MP) ratio model training. For each drug, the following variables are extracted or calculated:

- **MW, PSA, HBD, HBA, SMILES** : extracted from ChEMBL molecule API
- **Fu** : extracted from ChEMBL activity API (human plasma only)
- **LogP** : extracted from ChEMBL molecule API (alogp field)
- **pKa** : extracted from ChEMBL activity API (experimental values)
- **LogD7.4** : three columns: experimental (ChEMBL), calculated (from LogP + pKa), predicted (SMILES-based model)

PubChem is used as a fallback if a drug is not found in ChEMBL.



## Setup
```{r}
library(httr2)
library(jsonlite)
library(dplyr)
```



```{r}
drug_names <- c("Amphetamine", "Acetaminophen", "Aspirin", "Enalapril",
                "Ethinyl estradiol", "Ethosuximide", "Ivermectin", "Methohexital",
                "Norethindrone", "Phenobarbital", "Theophylline", "Thiopental",
                "Tinidazole", "Valproic Acid", "Zonisamide", "Alprazolam",
                "Amitriptyline", "Amlodipine", "Amoxapine")

cat("Total drugs:", length(drug_names))
```


## Step 1: Extract Molecule Properties from ChEMBL
Extracts MW, PSA, HBD, HBA, LogP, and SMILES from the ChEMBL molecule API.
Filters for exact name matches only. Duplicates are removed after joining.

```{r molecule-extraction, message=TRUE}
molecule_results <- data.frame()

for (drug in drug_names) {
  message(paste("Fetching molecule data for:", drug))
  
  url <- paste0("https://www.ebi.ac.uk/chembl/api/data/molecule/search.json?q=",
                utils::URLencode(drug), "&limit=10")
  
  req <- httr2::request(url) |> httr2::req_headers(Accept = "application/json")
  response <- httr2::req_perform(req)
  parsed <- jsonlite::fromJSON(httr2::resp_body_string(response), flatten = TRUE)
  
  if (is.null(parsed$molecules) || nrow(parsed$molecules) == 0) {
    message(paste("No results found for:", drug))
    next
  }
  
  match <- parsed$molecules |>
    dplyr::filter(tolower(pref_name) == tolower(drug)) |>
    dplyr::select(
      pref_name,
      molecule_chembl_id,
      molecule_properties.full_mwt,
      molecule_properties.psa,
      molecule_properties.hbd,
      molecule_properties.hba,
      molecule_properties.alogp,
      molecule_structures.canonical_smiles
    ) |>
    dplyr::slice(1)
  
  if (nrow(match) == 0) {
    message(paste("No exact match for:", drug))
    next
  }
  
  match$Drug_Input <- drug
  molecule_results <- dplyr::bind_rows(molecule_results, match)
}

molecule_results <- molecule_results |>
  dplyr::distinct(pref_name, molecule_chembl_id, .keep_all = TRUE)

cat("Drugs successfully matched:", nrow(molecule_results), "of", length(drug_names))
```

## Step 2: Extract Fu from ChEMBL Activity API
Extracts unbound fraction in plasma (Fu) from ChEMBL activity data.

Filtering logic:

- Human assays only: target_organism is NA or Homo sapiens
- Plasma measurements only: assay_description contains "plasma" or "human"
- Tissue measurements excluded
- Animal keywords excluded (using word boundaries to avoid substring matches)
- Fu values of exactly 0 or 1 are flagged for manual literature review



```{r fu-extraction, message=TRUE}
fu_results <- data.frame()

for (chembl_id in molecule_results$molecule_chembl_id) {
  message(paste("Fetching Fu data for:", chembl_id))
  
  tryCatch({
    url <- paste0("https://www.ebi.ac.uk/chembl/api/data/activity.json?molecule_chembl_id=",
                  chembl_id, "&standard_type=Fu&limit=100")
    
    req <- httr2::request(url) |> httr2::req_headers(Accept = "application/json")
    response <- httr2::req_perform(req)
    parsed <- jsonlite::fromJSON(httr2::resp_body_string(response), flatten = TRUE)
    
    activities <- parsed$activities
    
    if (is.null(activities) || !is.data.frame(activities) || nrow(activities) == 0) {
      fu_results <<- dplyr::bind_rows(fu_results,
        data.frame(molecule_chembl_id = chembl_id,
                   Fu_median = NA, N_fu = 0,
                   Fu_values = NA, Fu_flag = NA))
    } else {
      filtered <- activities |>
        dplyr::filter(
          (is.na(target_organism) | target_organism == "Homo sapiens") &
          !stringr::str_detect(tolower(assay_description),
                               "\\brat\\b|\\bmouse\\b|\\bmice\\b|\\banimal\\b|\\brabbit\\b|\\bdog\\b|\\bmonkey\\b") &
          !stringr::str_detect(tolower(assay_description), "tissue") &
          stringr::str_detect(tolower(assay_description), "plasma|human") &
          !is.na(standard_value)
        ) |>
        dplyr::mutate(standard_value = as.numeric(standard_value)) |>
        dplyr::filter(standard_value >= 0, standard_value <= 1)
      
      if (nrow(filtered) == 0) {
        fu_results <<- dplyr::bind_rows(fu_results,
          data.frame(molecule_chembl_id = chembl_id,
                     Fu_median = NA, N_fu = 0,
                     Fu_values = NA, Fu_flag = NA))
      } else {
        fu_summary <- filtered |>
          dplyr::summarise(
            Fu_median = median(standard_value, na.rm = TRUE),
            N_fu = dplyr::n(),
            Fu_values = paste(standard_value, collapse = ", "),
            Fu_flag = ifelse(any(standard_value == 0 | standard_value == 1),
                             "FLAG: contains 0 or 1", NA)
          ) |>
          dplyr::mutate(molecule_chembl_id = chembl_id)
        
        fu_results <<- dplyr::bind_rows(fu_results, fu_summary)
      }
    }
    
  }, error = function(e) {
    message(paste("Error processing", chembl_id, ":", e$message))
    fu_results <<- dplyr::bind_rows(fu_results,
      data.frame(molecule_chembl_id = chembl_id,
                 Fu_median = NA, N_fu = 0,
                 Fu_values = NA, Fu_flag = NA))
  })
}
```


## Step 3: Combine and Review
```{r combine}
final_dataset <- molecule_results |>
  dplyr::left_join(fu_results, by = "molecule_chembl_id") |>
  dplyr::rename(
    Drug      = pref_name,
    ChEMBL_ID = molecule_chembl_id,
    MW        = molecule_properties.full_mwt,
    PSA       = molecule_properties.psa,
    HBD       = molecule_properties.hbd,
    HBA       = molecule_properties.hba,
    LogP      = molecule_properties.alogp,
    SMILES    = molecule_structures.canonical_smiles
  ) |>
  dplyr::select(Drug, ChEMBL_ID, MW, PSA, HBD, HBA, LogP,
                Fu_median, N_fu, Fu_values, Fu_flag, SMILES)

final_dataset
```

## Step 4 : Quality Check

```{r quality-check}
cat("DATASET SUMMARY\n")
cat("Total drugs:", nrow(final_dataset), "\n")
cat("Drugs with Fu:", sum(!is.na(final_dataset$Fu_median)), "\n")
cat("Drugs missing Fu (need literature lookup):", sum(is.na(final_dataset$Fu_median)), "\n")
cat("Drugs with flagged Fu (need literature review):", sum(!is.na(final_dataset$Fu_flag)), "\n")
cat("Drugs missing MW (need investigation):", sum(is.na(final_dataset$MW)), "\n")

cat("\nDRUGS MISSING Fu\n")
final_dataset |>
  dplyr::filter(is.na(Fu_median)) |>
  dplyr::select(Drug, ChEMBL_ID) |>
  print()

cat("\nFLAGGED Fu VALUES\n")
final_dataset |>
  dplyr::filter(!is.na(Fu_flag)) |>
  dplyr::select(Drug, Fu_median, N_fu, Fu_values, Fu_flag) |>
  print()

cat("\nDRUGS MISSING PROPERTIES\n")
final_dataset |>
  dplyr::filter(is.na(MW)) |>
  dplyr::select(Drug, ChEMBL_ID) |>
  print()
```